### 2022-09-29 manuscript/manuscript.tex
more theory notes

Let us assume that there is a constant \emph{drift-effective} population size
$N_d$ which determines the rate of coalescence due only to neutral demographic
processes. Additionally, we assume some background selection process that in
expectation rescales the drift-effective population size by $b(t)$ at time $t$
(though this could be any process that modifies the coalescence rate). Then,
the expected pairwise coalescent time is

\begin{align}
  \E(T_2) = \sum_{i=1}^\infty \frac{i}{2 b(t) N_d}\prod_{t=1}^{i-1} \left(1-\frac{1}{2 b(t) N_d}\right).
\end{align}

The time-dependency of $b(t)$ is needed because coalescence rates essentially
speed up in the past under intermediate to weak selection, similar to
population growth \parencite{Walczak2012-fi,Seger2010-rs,Good2014-yz}.
Intuitively, this is because lineages are unlikely to coalesce with numerous
less fit lineages they encounter as they go back in time, until ultimately
coalescing in one of the least-loaded fitness classes.

Previous work in the strong background selection domain
\parencite{Charlesworth1993-gb,Nordborg1996-nq} approximates $\E(T_2)$ as $2
f_0 N_d$, where $f_0 = e^{-\nicefrac{U}{s}}$ is size of the least-loaded
fitness class. XXX on delay

Here, we model the background selection process forward in time. We approximate
the expected pairwise diversity with the probability of heterozygosity,
$\E(\pi) \approx \mathcal{H}$, which holds under weak mutation and random
mating, such that

\begin{align}
  %\mathcal{H} =  2 \mu \sum_{i=1}^\infty \prod_{t=1}^i \left(1-\frac{1}{2 b(t) N_d} \right)
  \mathcal{H} &= 1 - \sum_{i=1}^\infty (1-\mu)^{2i} \prod_{t=1}^i \left(1-\frac{1}{2 b(t) N_d} \right) \\
              &\approx 2 \mu \sum_{i=1}^\infty \prod_{t=1}^i \left(1-\frac{1}{2 b(t) N_d} \right)
\end{align}

in the limit where $\mu \to 0$. This is equation 4 of
\textcite{Santiago2016-mu}; they define the sum as the heterozygosity effective
population size.

\subsection*{Forward in Time Models of Background Selection}

Next, show how quantitative genetic models of linked selection
\parencite{Robertson1961-ho,Santiago1995-hx,Santiago1998-bs,Santiago2016-mu}
model $b(t)$ forward in time.

\begin{align}
  b(t) = \frac{N_v}{N_d}
\end{align}

where

>>>>

Let us assume that there is a constant \emph{drift-effective} population size
$N_d$ which determines the rate of coalescence due only to neutral demographic
processes. Additionally, we assume some background selection process that in
expectation scales the effective population size by $b(t)$ at time $t$ (though
this could in theory be due to any process). Then, the expected pairwise
coalescent time is

\begin{align}
  \label{eq:T_2}
  \E(T_2) = \sum_{i=1}^\infty \frac{i}{2 b(t) N_d}\prod_{t=1}^{i-1} \left(1-\frac{1}{2 b(t) N_d}\right).
\end{align}

The time-dependency of $b(t)$ is required because the background selection
process will usually alter the rate of coalesce through time. For example,
under weak and intermediate selection, coalesce rates speed up as lineages move
in the past under intermediate to weak selection, similar to population growth
\parencite{Walczak2012-fi,Seger2010-rs,Good2014-yz}.

% Intuitively, this is
% because lineages are unlikely to coalesce with numerous less fit lineages they
% encounter as they go back in time, until ultimately coalescing in one of the
% least-loaded fitness classes.

This formulation is general and classic background selection theory (in the
no-ratchet domain) is a special case of it. Under classic BGS and load theory,
two randomly sampled lineages will each carry a number of deleterious mutations
sampled from a Poisson stationary distribution. Because any lineage carrying $j
> 0$ strongly deleterious mutations is destined to extinction, these lineages
can only coalesce in a common ancestor in the least-loaded class with zero
deleterious mutations. Under the Poisson stationary distribution, the size of
this class is $f_0 N$, where $f_0 = e^{-\nicefrac{U}{s}}$, so the coalescent
rate is scaled by $\nicefrac{1}{f_0}$. Once the lineages reach this fitness
class, there are no fitness differences among individuals, so they coalesce
following the Kingman process. A lineage moves backwards in time $t_d$
generations through the ``delay phase" to the least-loaded class. Classic BGS
theory ignores this delay phase, since $t_d \sim \log(j+1)/s << f_0 N$
(Durrett, p. 213).

%Here, we model the background selection process forward in time. We approximate
%the expected pairwise diversity with the probability of heterozygosity,
%$\E(\pi) \approx \mathcal{H}$, which holds under weak mutation and random
%mating, such that

%\begin{align}
%  %\mathcal{H} =  2 \mu \sum_{i=1}^\infty \prod_{t=1}^i \left(1-\frac{1}{2 b(t) N_d} \right)
%  \mathcal{H} &= 1 - \sum_{i=1}^\infty (1-\mu)^{2i} \prod_{t=1}^i \left(1-\frac{1}{2 b(t) N_d} \right) \\
%              &\approx 2 \mu \sum_{i=1}^\infty \prod_{t=1}^i \left(1-\frac{1}{2 b(t) N_d} \right)
%\end{align}

%in the limit where $\mu \to 0$. This is equation 4 of
%\textcite{Santiago2016-mu}; they define the sum as the heterozygosity effective
%population size.

\subsection*{Forward in Time Models of Linked Selection}

Next, show how quantitative genetic models of linked selection
\parencite{Robertson1961-ho,Santiago1995-hx,Santiago1998-bs,Santiago2016-mu}
essentially model $b(t)$ with an analogous forward in time process. Their
approach is to consider how the variance in neutral allele frequency change is
impacted by a linked selection process. Under a Wright--Fisher model, the
variance in allele frequency change is also determined by a pairwise
coalescence rate, often expressed as a \emph{variance-effective} population
size,

\begin{align}
  \V(p_t) = p_0(1-p_0)\left(1 - \prod_{i=1}^t \left(1-\frac{1}{2N_e(t)}\right)\right)
\end{align}

where if $N_e(t) = N_e$ and $t=1$, this is the familiar Wright--Fisher variance
$\V(p_1) = p_0(1-p_0) / 2N_e$. When selection is not acting, allele frequency
changes per generation are independent, and neutral diversity is set by this
single $N_e$.

XXX

Since we track an arbitrary neutral allele's frequency $p_i$, by symmetry,
linked selection has no impact on the direction of change, $\E(\delta p_i) =
0$. We derive an expression for $\V(p_t)$ by induction following the approach
of \textcite{Santiago1995-hx}. In the first generation, the frequency change
can be partitioned into three components. Conditioning on the parent's
heritable fitnesses, we can imagine constructing an infinite number of
evolutionary replicates. Across all replicates, a random neutral allele's is
pushed

\begin{align}
  p_1 - p_0 = \sum_{i=1}^{N} 
\end{align}

Since the variance in neutral allele frequency change is also determined by the
pairwise coalescence rate, this provides us with a way to model $b(t)$ under
linked selection.

These models derive an expression for the variance-effective population size of
a focal neutral site, $N_v(t)$, based on how linked fitness variation perturbs
the frequency trajectory at this site.  The variance-effective population size
for a constant $N_e$ is defined by the variance in allele frequency change at
the neutral site, over evolutionary replicates,

where if $t=1$, this is the familiar Wright--Fisher variance $\V(p_1) =
p_0(1-p_0) / 2N_e$. When selection is not acting, allele frequency changes per
generation are independent, and neutral diversity is set by this
single $N_e$.

However when selection is acting at linked sites, neutral diversity is
determined by the long-run $\V(p_t)$ as $t \to \infty$. Intuitively, this is
because the rate of drift is magnified by autocorrelation in allele frequency
change (XXX), and diversity is determined by the cumulative impact of these
autocorrelations in the long-run. This autocorrelation effectively reduces the
coalescence rate in a time-dependent manner, analogous to equation
\eqref{eq:T_2}. Thus, a single constant $N_e$ is insufficient to describe
levels of neutral variation, since it is insufficient

The variance in allele frequency change after $t$ generations is

\begin{align}
  \V(p_t) = p_0(1-p_0)\left(1 - \prod_{i=1}^t \left(1-\frac{1}{2N_e(t)}\right)\right)
\end{align}

This is the basis of the quantitative genetic models of linked selection: a
forward in time model for $\V(p_t)$ implies a pairwise coalescence rate
function $b(t)$, which determines statistics based on pairwise coalescence such
as $\E(\pi)$.

Next, we consider how fitness variation impacts $\V(p_t)$.

Throughout, we will take the variance $\V(p_t)$ to be the variance over
evolutionary replicates. Since the variance is for a random neutral allele's
frequency,

a reasonable approximation to how background selection process
rescales the coalescent rate at this site is given by $b(t) =
\nicefrac{N_v(t)}{N_d}$.
